### 2015-07-02 doc/mec-14-1243-response-letter.md

Dear Dr. Hoffmann,

We appreciate the comments from you and the reviewers. Both reviewers raise important points and concerns, to which we have responded below with changes to the manuscript and/or rebuttal if we disagreed with the reviewer’s suggestion. We feel that this process has helped us to build a stronger case both for the methodology and main results of our study. We especially thank the reviewers for highlighting ambiguities in some of our original language.  While we respond to the reviewers comments in details below, we address the three major concerns here.

1) Acclimation vs adaptation

Reviewer 1 is concerned that we “cannot tease apart genetic or plastic differences in expression profiles” due to using field-collected ant colonies in this experiment. While a complete-generation common garden experiment is preferable for distinguishing genetic from plastic differences, this is neither possible nor practical for the vast majority of species, including Aphaenogaster. Moreover, rearing a single generation in common conditions does not resolve all these issues (e.g. maternal genetic effects). A key outcome of our work is demonstrating a method to gain insight on the relative role of genetic or plastic differences even when rearing organisms in common gardens is not feasible. We limited differences in developmental acclimation by maintaining colonies in the lab for 6 months prior to sampling. Given that a new cohort is reared in 2-3 months, 6 months allowed for a substantial number of workers to be fully reared under common-garden conditions, such that our samples would have contained a mix of truly common-garden and adult-acclimated workers. This is actually better than has been ascribed to adaptation and published for other long-lived taxa including corals (Barshis et al. 2013 PNAS 110:1387) and fish (Smith et al. BMC Genomics 2013, 14:375). With hindsight, the reviewer’s suggestion of sampling the species from sites where they overlap is great, but unfortunately was not done. However, it is worth noting that in previous work where the species were collected from overlapping sites along an altitudinal gradient in the Smokies (Warren and Chick 2013), both CTmax and CTmin differed significantly, suggesting different underlying physiological adaptations such as we have described here.

We have addressed this concern, noting the points above, in the section “Common Garden Design” in the Methods (lns 165 - 177).

2) Replication of RNA-seq

The standard practice of replication in RNA-seq experiments was developed for ANOVA to identify differences in expression among tissues and discrete conditions (e.g. disease states). For these studies, replication is absolutely essential. In contrast, ecologists are often interested in the response across a gradient, which is more directly modeled by a linear regression analysis (Cottingham et al. 2005). Variance in the expression profiles is estimated from the linear regression model. Replication at any particular temperature is not necessary, as power to detect a relationship is most improved by adding additional temperatures, not by replicating those that have already been assessed.  In fact, this approach results in a more precise variance estimate than if we had replicated only a few temperatures (Cottingham et al 2005). Only transcripts with a significant linear or quadratic relationship with temperature, after correcting for false discovery rate, were retained in the analysis.  Such analyses are just beginning to be used in a transcriptomic context (the only other example we are aware of is Hodgins-Davis et al 2012 GBE). To better clarify our approach for readers, we have added an expanded justification for the regression design in the Introduction  (lns 113-116):

To characterize the thermal reactionome, we characterized the reaction norm for each gene using a regression-based statistical approach to identify temperature-dependent patterns of change in gene expression. We used these response patterns to quantitatively test three mechanistic hypotheses of thermal adaptation.

and explanation in the methods  (lns 178-183):

	Unlike ANOVA-based experimental designs, which derive statistical power from replication within each experimental treatment level, regression designs have greater power when sampling additional values across the range of the continuous predictor variable (Cottingham et al 2005). Thus, we focused our sequencing efforts on maximizing the number of temperatures at which the transcriptome was profiled, rather than on replication at each temperature.

2) Validation with qPCR

This is an important issue to which we gave careful consideration. For two reasons, we did not feel that the results of qPCR would improve this study.

First, RNA-seq is no longer a new methodology. Many studies have been published in ecological journals using RNAseq data without qPCR validation, such as Barshis et al. (2012 PNAS) and Smith et al. (2013 BMC Genomics).

Second, it is only feasible to perform qPCR on a small subset of transcripts. In our approach, we avoid focusing on the response of individual transcripts (e.g we *do not* provide a list of differentially-expressed genes). Instead, we focus on transcriptome-wide responses to temperature, which groups large numbers of genes into distinct response patterns. While our results are undoubtedly wrong for some individual transcripts, they are likely to be distributed at random in regards to these classifications. Thus, while qPCR may prove that the expression patterns of some transcripts are incorrect, it is unlikely to change the results for whole functional groups.

> Editor Comments to Author:
> 
> This interesting paper has received three careful reviews. The first reviewer recommended rejection, the second was enthusiastic about the paper, and the third sat somewhere in the middle. The first reviewer points out that the work does not tightly control for genetic versus plastic effects and therefore the paper does not precisely test the issues of interest to the authors. I suspect that this can be fixed with reworking of the paper in terms of interpretations, taking care to separate out the different components where appropriate. The reviewer also makes some important points about replication and the measurement of CTmax as well as the nature of control thermal treatments. The second reviewer was complimentary about the overall approach and had only minor comments, which cover some confusion stemming from the text and one of the Figures. Some apparent errors in the wording are also noted. The third reviewer also raises the issue of replication at the different temperatures and has an issue with the terminology in the paper and division into a predefined set of categories. The comments are worth taking on board. The reviewer also points out that some aspects of expression changes are likely to be missed, and calls for new qPCR data as is often carried out. The reviewer points out where results are inconsistent with the literature and novel patterns that appear not to be consistent with expectations, which all need discussion. The authors should take these comments on board when making interpretations of their data, and when interpreting the patterns established from their work. The issue of replication needs consideration along with measurement error and validation of expression patterns established from RNAseq.
> 
> Dr. Ary Hoffmann
> Subject Editor, Molecular Ecology

> Reviewer Comments to Author:
> Reviewer: 1
> 
> Recommendation: Reject
> 
> Comments:
> 
> The current paper by Stanton-Geddes et al generates “reactionomes” by generating RNA seq data  for two different species of ant at 12 different temperatures (0-38.5C).  The authors estimate CTmax in both species and link changes in their expression profiles to these estimates of CTmax and the environments where the species were collected. I think the authors were ambitious, the questions they present relevant and highly topical however, I believe there are significant design issues which limit the interpretation of the results.
> 
> First and fore mostly, if I understand the methods correctly, ants were collected from the field acclimated at 25C for 6 months and then experiments performed. While I understand that many species can be difficult to rear in the laboratory. Without performing a true common garden experiment (i.e both species of ants developed and reared under standard conditions) you simply cannot tease apart genetic or plastic differences in expression profiles. Are the observed differences between the species due to evolved differences related to thermal evolution or simply a consequence of developmental acclimation, by collecting cold adapted species from a cold environment and vice versa only confounds the results even more. One possible way that the authors may have alleviated this issue, to some degree, is to choose 2 species that vary in their distribution but where their distribution overlaps. This may be the case for these two species but the authors have not made that clear in their ms. If their species do overlap in distribution the authors should at the very least have collected these species from the same site. With such a design you could have then assumed that developmental acclimation may have been similar across the two species. Albeit you would still have GxE effects and these are likely to differ between species.

We address this concern as (1) above.

> My other major concern, which also depends critically on whether I have understood the methods correctly (see below comments) is that for the RNA seq the authors have pooled 3 ants per treatment (not a particular high number) and actually have no replicates per treatment. How do you estimate the variance in your expression profiles?

We address this concern as (2) above.

> Minor comments
> 
> -Although authors have characterized CTmax in the two species, how do these estimates of CTmax relate to the heat treatments that the authors use? Have the authors really examined a phenotype here? The authors CTmax estimates will also be confounded by developmental acclimation.
>

The maximum temperature we heated ants to was 38.5℃, below the CTmax for both A. carolinensis (~43℃) and A. picea (~42℃) (Fig 3. Warren & Chick 2013), so we have confidence that gene expression changes reflected active responses to thermal stress rather than dysregulation caused by loss of function. We have added this information to the “Common Garden Design” sub-section of the Methods  (lns 191 - 194).

Both colonies were maintained in the lab for 6 months prior to experimentation, so differences in developmental acclimation are as minimized as the biology of the species will allow, but it is of course true that we cannot rule out some residual developmental effect. To reflect this remaining potential source of phenotypic plasticity, we have attempted throughout the ms to make our word choices more precise in regard to whether we have demonstrated evidence for evolutionary differences.

> -The authors need to clearer about where in the distribution of the ant species they are sampling. Are they sampling them from their extremes? Middle?

We have clarified this in the Methods  (lns 154 - 155).

    These locations are centrally located within each species’ geographic range.

> 
> -Line 40-55 The end of this paragraph reads like methods
> 
> -Line 56-60 also reads like methods
>

We appreciate the reviewer’s concern, and removed some of the detailed information (lines 56-57 from draft 1), but we do feel feel that some methodological details are necessary here to explain how we tested our hypotheses.  We have attempted to reframe these such that the reason for their inclusion is more clear.

> -Line 57 it would be helpful to have a figure of the temperatures you used here, or at least have it written out in the methods
> 
> -I found the methods section confusing and not detailed enough
> 
> -Line 97 makes it sound like you collected each ant species form Molly Bog Vermont, but I don’t think that’s  what you mean I think you mean you collected a single colony of A. picea from Molly bog and a single population of A. carolinensis from Durham North Carolina

True - we have clarified this in the text  (lns 150 - 153):

	In fall 2012, we collected a single colony of Aphaenogaster picea from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508° N, -72.702° W) and a single colony of Aphaenogaster carolinensis, part of the A. rudis species complex (Umphrey 1996), from Durham, North Carolina (36.037° N -78.874° W).

> 
> -Why were you not able to determine the CTmax from the original,

It would have been ideal to have paired measures such as these, but both transcriptome and CTmax measurements require a large number of workers, and the colonies were simply not large enough to do both. We’ve added this information to ln 157 of the manuscript.

could you please describe your CTmax methods, there are many different ways this is calculated. It shouldn’t be necessary for me to have to look up another paper to understand your methods.

We have added this information as requested  (lns 160 - 163):

	We tested the upper and lower critical thermal limits for 5 ants from each of these colonies using a ramp of 1° C per minute, starting at 30° C, and recorded the temperature at which the ants were no longer able to right themselves (Warren and Chick 2013).

> 
> -Line 109 To be clearer could the authors state that this part of the experiment was performed on ants collected in 2012

Done  (ln 150).

> 
> -The sampling and replicates are not clear Line 109-120. The authors collected 10 ants from each colony and exposed them to a single random temperature on a given day for a period of 12 days. Ants were then pooled into 3 which would make me think you had 10 ants per species per treatment you pooled them into 3 to make almost 4 replicates but on line 122 you say for each species the 12 samples were barcoded?

We apologize for the confusion. We collected 10 ants so that we would have extra samples if necessary, though only 3 were used in pooling (per temperature) for sequencing. For clarity, we now simply refer to 3 ants.

> 
> -Did the authors ensure these treatments did not cause mortality?

Yes. We have added this to the text (lns 191 - 193).

> 
> -Should the authors have a control here i.e 24.5 and look at species specific expression changes?
>

It’s unclear what the reviewer means here. A major advantage of the regression approach is that we have quantitative information on transcript expression in response to temperature and can predict expression at 24.5, or any other temperature, from the model of each transcripts’ reaction norm.  Species-specific expression changes are identified by a significant species x temperature interaction.  We have adopted the reviewer’s suggestion below to use 24.5 as our baseline value (see response to comment for line 222).

> -Line 162 I am unsure what the authors are suggesting here, that the data was odd so the got rid of it?

As you can see in this figure, which we now include in the Supporting Information, the “Ar_7” (Ar = A. carolinensis) sample clusters with all the other A22 (= A. picea) samples, while the “A22_7” sample clusters with all the other Ar samples. The misplaced samples are marked with red arrows. The most parsimonious explanation for this is a labeling mistake. Removing these samples, rather than relabeling, is the conservative approach.

> 
> -Line 169 Rather than writing out formula could you at least include a description of your model. Where species considered random or fixed effects? You can run mixed model glm’s in R and maybe this is what you did but your description is unclear. It is also unclear to me what temp and temp2 are in the model, this is not explained anywhere .
>

The formatting of this equation was lost in the file conversion, temp2 should have been temp2 indicating that we were fitting a second-degree polynomial regression. All effects were fixed as temperature is continuous and we only had two species. We have clarified in the text (lns 250 - 255):

	Temperature and species were both fixed effects, with a quadratic term included for temperature.

> -Line 180-184 I am unsure what exactly you are performing the step wise regression on?

The statistical significance of the species and linear and quadratic temperature terms in the model were tested via step-wise model selection.

> -Line 188-192 What about down regulated transcripts?

In a reaction norm context, downregulation is the inverse of up-regulation. That is, transcripts downregulated at low temperatures fall into the *High* category, transcripts downregulated at high temperatures fall into the *Low* category.  Splitting the temperature treatments into two at the midpoint to consider only low or only high temperatures would potentially have allowed us to make distinctions between upregulation and downregulation in these cases, but with only 12 temperatures in total, we would unfortunately be left with too few temperatures in each case to estimate the model.  We have clarified this in the sub-section “Identification of thermally-responsive transcripts”  (lns 279 - 284)

Because expression category was defined by the temperature of maximal expression, both Low and High categories were biased toward transcripts up-regulated at that temperature extreme, but also likely included some transcripts down-regulated at the opposing extreme. The two categories which could unambiguously distinguish up- from down-regulation are Bimodal (up at both extremes) and Intermediate (down at both extremes).

> 
> -The authors grouped their transcripts by expression level and temperature, the temperature categories need better justification

The temperature categories were decided a priori to allow us to apply the results to the biological hypotheses we were testing. This is elaborated in the sub-section “Identification of thermally-responsive transcripts”  (lns 271 - 284):

	We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined a priori to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species...

> 
> -Line 222 It seems odd that the authors have used 19.5C as their baseline for constitutive expression when ants were reared for 6 months at 25C. It seem 24.5 C would be a more suitable control.

This is a good point that we had not considered.  As suggested by the reviewer, we changed the baseline for constitutive expression to the rearing temperature of 25℃. See commit afdd69d on github (https://github.com/johnstantongeddes/ApTranscriptome). We use this version of the results in the revised manuscript, though it has no qualitative effect on the major results.

> 
> -Line 322 the authors start talking about down regulation but as I understood their categories high was high expression at high temperatures, low was high expression at low temperatures etc. Maybe I have misunderstood something here?

See our comment above.

> 
> -Line 424 It is unclear what the authors are trying to say here are they suggesting you can link the number of responsive genes to the level of additive genetic variation?
>

We were presenting a post-hoc explanation for the observation that many more genes were responsive to cold than warm temperature extremes. As we emphasized in the text (e.g. To the extent that …) this was a speculative explanation for this observation, though consistent with other work (Kellerman et al 2012, Warren and Chick 2013). However, given the uncertainty about the explanation we have chose to remove this section from the manuscript.

> 
> 
> Reviewer: 2
> 
> The authors of this study characterize the thermal reaction norms of all genes in a de novo transcriptome assembly across 12 temperatures spanning the entire thermal breadth of two species of ants: the cool-climate Aphaenogaster picea and the warm-climate Aphaenogaster carolinensis. In the past, studies exploring transcriptomic responses to temperature have mostly used ANOVA-type experimental designs, with only a few temperature levels. But continuous variation is much better characterized with the reaction norm approach used in this study, describing variation among genes and among organisms in the shape of expression responses across temperature. I really enjoyed this paper -- the approach taken here is exciting, and I found the manuscript to be quite well written, with clearly defined hypotheses an expectations. The comparative reaction norm approach taken here is, to my knowledge, unprecedented, and will make an important contribution to the field of thermal biology. I strongly recommend this manuscript for publication (and look forward to its acceptance so I can cite it!). I have only a few minor comments and revisions, outlined below.
>

We really appreciate these comments. While this study isn't perfect, we also believe that this study is pushing back the boundaries of our knowledge of thermal biology.

> Lines 61-86. As the authors note, studies exploring transcriptomic responses to stress have sometimes been prone to storytelling, so I appreciate the specific predictions generated by each of the three potential mechanisms of stress adaptation.
>

This problem of storytelling is rife in the ecological genetic literature, and one of our motivations for writing this paper was to demonstrate how similar methods can be analyzed in an alternate approach to gain novel insight.

> Figure 1
> 
> -Figure 1 is key to the whole paper, but connecting the results to the predictions generated by each of the three hypotheses made my head hurt a little. I wonder if there is a way to make it easier for the reader to draw inferences from this figure? One possibility would be an inset cartoon showing the results that could be expected in support of each of the competing hypotheses? Also maybe include little line cartoons along the axes depicting each of the norms of reaction.
>

This is a useful suggestion and we have made a figure (Fig. 1) with illustrations of the expected results under the *enhanced response* and the *tolerance* hypotheses that can be compared to the true results.

> The categories “High” “Low” “Intermediate” and “Bimodal” are defined well in the manuscript, but it’s always nice when key figures can stand alone without reference to the text.
>

We define these in the caption to Table 1, and refer to this in Figure 1 rather than repeating. We could also switch this if preferred.

> - What are the units in the color scale?

The units are number of transcripts. We have updated the legend to clarify this:

	...the color of each cell (purple = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts.

> 
> 196-197. I’m missing something about the methods - lines 196-197 state that the analyses focused on transcripts with a significant temperature x species interaction but in Figure S1, 4/5 of the boxes on the diagonal are blue (= excess of transcripts in the same reaction norm category for both species). Shouldn’t these have been excluded by focusing only on transcripts where temperature effects differed between species?
>

We apologize for being unclear in the text. For the analyses of the how the number of responsive transcripts differs between species (first paragraph under “Statistical Analyses”), we used all responsive transcripts, both those with a significant temperature x species interaction and those with a main effect of temperature only. The justification for including all thermally-responsive transcripts is that the enhanced response hypothesis explicitly predicts that the genes upregulated by the species that sees the stressor infrequently should retain that function in the other, while under the tolerance hypothesis they should not.

In contrast, for the comparisons of the *temperature of gene activation* (second paragraph under “Statistical Analyses”), only transcripts with a significant temperature x species interaction are used as, by definition, there is no difference in expression levels between species 
We have changed the text (lns 292-295) to clarify this.

### 2015-07-02 ApTranscriptome_TR.Rmd
Final edits for response. Added response letter

---
output: word_document
---
Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
========================================================================================

@@

**October 14, 2014**

**Technical Report No. 3**

@@

```{r setup, echo=FALSE, results='hide', message = FALSE}
# Global settings
options(stringsAsFactors=FALSE, digits = 3, scipen = 1)

@@

In this technical report, which accompanies the manuscript **Patterns of stress tolerance versus resistance in genome-wide expression data of parapatric ant species** (Stanton-Geddes et al., in press), we:

1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
2. identify thermally-responsive genes
3. evaluate differences in reaction norms of expression for each thermally-responsive gene between the two species
4. perform gene set enrichment analysis of thermally-responsive genes for the two species

This script is completely reproducible assuming that R, `knitr` and the other required libraries (listed within the source document) are installed on a standard linux system using the following:

    Rscript -e "library(knitr); knit('ApTranscriptome_TR.Rmd')"

@@

The raw Illumina fastq files are available from the NCBI short read archive [link tbd]. The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*. For historical reasons, I sometimes refer to these species as colonies at times throughout this technical report.

For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5C ranging from 0C to 38.5C, for one hour in glass tubes in a water bath. The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen). Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20e6 and 16e6 reads for the A22 and Ar samples, respectively.

@@

~~~
java -jar trimmomatic-0.30.jar PE -threads 40 -phred33 -trimlog trimmomatic.log sample.R1.fastq sample.R2.fastq sample.R1.trimmed.paired.fastq sample.R1.trimmed.unpaired.fastq sample.R2.trimmed.paired.fastq sample.R2.trimmed.unpaired.fastq ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:36
~~~

@@

Properly paired and unpaired reads passing the Trimmomatic filter were combined and used in de novo transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`. Unpaired reads resulting from Trimmomatic were concatenated to the set of left (R1) reads per Trinity usage documentation. Assembly on a single 40 CPU machine with 1TB of memory required 20 hours of compute time.

~~~
Trinity.pl --seqType=fq --JM=90G —left=all.R1.trimmed.fastq —right=all.R2.trimmed.fastq --output=trinity --CPU=40 --inchworm_cpu=40 --bflyCPU=5
~~~

This assembly contained 100,381 unique components (roughly genes) in 126,172 total transcripts (Table 1).

@@

```{r assemstats, results = "asis", echo=FALSE}
# make table
trinity <- c("126,172", "100,389,539", "358", "795", "16,201", "1,631")
reduced <- c("105,536", "62,648,997", "320", "593", "15,491", "895")

assemstats <- rbind(trinity, reduced)
colnames(assemstats) <- c("Total contigs", "Total length", "Median contig size", "Mean contig size", "N50 contig", "N50 Length")

@@

    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust -dbs hsref,bast,vir,arch

@@

Running Trinity and subsequent programs is time and memory-intensive so the final assembly can be downloaded and used for all further analyses. In addition, the compressed archive contains the "clean" and "contaminant" sequences after filtering with DeconSeq.

~~~
# download filtered Trinity assembly, uncompress and move
wget http://johnstantongeddes.org/assets/files/ApTranscriptome/Aphaenogaster_transcriptome.tar
# check md5sum
md5sum Aphaenogaster_transcriptome.tar
# fd6dbb0b3e88e1c797c9e74611b245b2

# move and extract
mkdir -p results/
mkdir -p results/trinity-full/
mv Aphaenogaster_transcriptome.tar results/trinity-full/.
tar -xvf Aphaenogaster_transcriptome.tar
~~~

@@

Results are available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).

This annotation file can be read directly to R:

```{r download, echo = TRUE}
# URL for annotation file
annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome/ApTranscriptome_AnnotationTable_20140113.txt")
# load 
annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotation.file)

@@

### Quantify gene expression ###

@@

~~~
# download gene expression quantification
wget http://johnstantongeddes.org/assets/files/ApTranscriptome/sailfish_quant_20140916.tar
# check md5sum
md5sum sailfish_quant_20140916.tar
# 26102c7ef86cf30b5f8e923640378185

@@

@@

### Grouping of thermally-responsive transcripts

The set of transcripts with significant expression patterns include those with expression that differs by species, temperature and the interaction of species and temperature. In this section, I am specifically interested in the thermally-responsive transcripts (temperature and species x temperature) so I subset the significant transcripts to examine these.

```{r responsive_transcripts}
interaction.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
other.lms <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.lms))]
temperature.lms <- other.lms[which(Map(grepFunc, other.lms, term = "val") == TRUE)]
colony.lms <- other.lms[setdiff(names(other.lms), names(temperature.lms))]
responsive.lms <- c(temperature.lms, interaction.lms)
rm(other.lms)

@@

### Thermal-response functional types ###

The previous section identified the transcripts with thermally-responsive expression. In this section, I determine the shape of the expression response to temperature for each transcript. Catego
ries of expression response are:

* High - increase expression with temperature
* Low - decrease expression with temperature
* Intermediate - maximum expression at intermediate temperatures (14 - 28C)
* Bimodal - expressed greater than one standard deviation of expression at both low and high temperatures

I do this first for the thermally-responsive transcripts where there is no interaction with species. For the transcripts where thermal-responsive expression depends on species, I determine the functional type of the expression response separately for each species.

@@

Next, I compare the number of thermally-responsive in each response category between the two colonies.

@@

@@

To identify specific cells that deviate from null expectation, calculate expected observations assuming that marginal frequencies equal the overall frequency for each expression category, then determine which cells have deviations between observed and expected that are greater than the overall X^2^ statistic.

@@

To visualize this, make a heatmap that shows deviation from expected count.

@@

Per suggestion of a reviewer, make a cartoon figure that shows the contrasting results of the *enhance defense* and the *tolerance* hypotheses.

The *enhanced response hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response. This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g. more shared "Low" transcripts) and a deficit of transcripts that shift to other response patterns or become not responsive.

>>>>

---
output: html_document
---

Technical Report for *Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species*
========================================================================================

@@

**June 24, 2015**

@@

```{r setup, echo=FALSE, results='hide', message = FALSE, warning = FALSE}
# Global settings
options(stringsAsFactors=FALSE, digits = 3, scipen = 1)
options(width = 80)

@@

In this technical report, which accompanies the manuscript **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**, we:

1. Describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`.
2. Measure the reaction norm across a thermal gradient (0°C to 38.5°C) for all transcripts and characterize the *thermal reactionome* as the set of transcripts that are thermally-responsive.
3. Quantitatively test three mechanistic hypotheses of thermal adapation:.
  - The *Enhanced response* hypothesis proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
  - The *Tolerance hypothesis* proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency.
  - The *Genetic assimilation hypothesis* proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.

We quantify the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species. We use gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.

@@

The raw Illumina fastq files are available from the NCBI short read archive [link tbd]. The assembled transcriptome, annotation and expression values are available from the Harvard Forest Archives, HF-113 [http://harvardforest.fas.harvard.edu/data-archive] datasets [hf113-36](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-36-ap-ann-table.csv), [hf113-40](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-40-ap-trans.tar) and [hf113-41](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-gene-exp-quant.tgz). These processed files are used for this analysis due to the computational demands of re-generating the files, but the exact commands for each of these steps are documented below.

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*.

For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5°C ranging from 0°C to 38.5°C, for one hour in glass tubes in a water bath. The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen). Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20e6 and 16e6 reads for the A22 and Ar samples, respectively.

@@

~~~
java -jar trimmomatic-0.30.jar PE -threads 40 -phred33 -trimlog trimmomatic.log \\
sample.R1.fastq sample.R2.fastq sample.R1.trimmed.paired.fastq sample.R1.trimmed \\
.unpaired.fastq sample.R2.trimmed.paired.fastq sample.R2.trimmed.unpaired.fastq \\
ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:36
~~~

@@

Properly paired and unpaired reads passing the Trimmomatic filter were combined and used in de novo transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`. Unpaired reads resulting from Trimmomatic were concatenated to the set of left (R1) reads per Trinity usage documentation. Assembly on a single 40 CPU machine with 1TB of memory required about 20 hours of compute time.

~~~
Trinity.pl --seqType=fq --JM=90G —left=all.R1.trimmed.fastq —right=all.R2.trimmed.fastq \\
--output=trinity --CPU=40 --inchworm_cpu=40 --bflyCPU=5
~~~

This assembly contained 100,381 unique components (~genes) in 126,172 total transcripts (Table 1).

@@

```{r assemstats, results = "asis", echo=FALSE}
# make table
Trinity <- c("126,172", "100,389,539", "358", "795", "16,201", "1,631")
Reduced <- c("105,536", "62,648,997", "320", "593", "15,491", "895")

assemstats <- rbind(Trinity, Reduced)
colnames(assemstats) <- c("Total contigs", "Total length", "Median contig size", "Mean contig size", "N50 contig", "N50 Length")

@@

    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust \\
    -dbs hsref,bast,vir,arch

@@

Running Trinity and subsequent programs is time and memory-intensive so the final assembly is downloaded and used for all further analyses. The downloaded archive contains the "clean" and "contaminant" sequences after filtering with DeconSeq.

~~~
# download filtered Trinity assembly, uncompress and move
wget http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-40-ap-trans.tar

# move and extract
mkdir -p results/
mkdir -p results/trinity-full/
mv hf113-40-ap-trans.tar results/trinity-full/.
tar -xvf hf113-40-ap-trans.tar
~~~

@@

Results were available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).

This annotation file can be downloaded from the Harvard Forest [archive](http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-36-ap-ann-table.csv), or read directly into R.

```{r download, echo = TRUE}
# URL for annotation file
annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome/ApTranscriptome_AnnotationTable_20140113.txt")

# load 
annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotation.file)

@@

### Gene expression ###

@@

~~~
# download gene expression quantification
wget http://harvardforest.fas.harvard.edu/data/p11/hf113/hf113-41-ap-gene-exp-quant.tgz
# http://johnstantongeddes.org/assets/files/ApTranscriptome/sailfish_quant_20140916.tar
# check md5sum
md5sum sailfish_quant_20140916.tar
# 26102c7ef86cf30b5f8e923640378185

@@

\newpage

@@

The set of transcripts with significant expression patterns include those with expression that differs by species, temperature and the interaction of species and temperature. I am specifically interested in the thermally-responsive transcripts (temperature and species x temperature) so I subset the significant transcripts to examine these.

```{r responsive_transcripts}
interaction.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
other.lms <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.lms))]
temperature.lms <- other.lms[which(Map(grepFunc, other.lms, term = "val") == TRUE)]
colony.lms <- other.lms[setdiff(names(other.lms), names(temperature.lms))]

# combine all responsive transcripts
responsive.lms <- c(temperature.lms, interaction.lms)
rm(other.lms)

@@

I checked the proportion of responsive transcripts against those that have a BLAST hit.

```{r responsive_with_BLAST}
annotation.table.responsive <- annotation.table[which(annotation.table$best.hit.to.nr != "-"), ]

responsive_transcripts_w_BLAST <- length(which(unique(resp.TPM.dt.sub$Transcript) %in% annotation.table.responsive$Sequence.Name))

responsive_transcripts_w_BLAST / round(nrow(annotation.table.responsive))
```

### Thermal-response expression categories ###

The previous section identified the transcripts with thermally-responsive expression. In this section, I determined the shape of the expression response to temperature for each transcript. Categories of expression response are:

- **High** - increase expression with temperature
- **Low** - decrease expression with temperature
- **Intermediate** - maximum expression at intermediate temperatures (14 - 28C)
- **Bimodal** - expressed greater than one standard deviation of expression at both low and high temperatures
- **Not Responsive** - not thermally-responsive in one species but does respond in the other

For the transcripts where thermal-responsive expression depends on species, I determined the functional type of the expression response separately for each species.

@@

Next, I compared the number of thermally-responsive in each response category between the two colonies.

@@

# proportion
Ap.type.table.prop <- round(Ap.type.table / (sum(Ap.type.table)/2) * 100)

@@

To identify specific cells that deviate from null expectation, I calculated the expected observations assuming that marginal frequencies equal the overall frequency for each expression category, then determined which cells have deviations between observed and expected that are greater than the overall X^2^ statistic.

@@

To visualize this, I made a heatmap that showed the deviation of the actual number of transcripts in each cell from the expected count.

@@

The *enhanced response hypothesis* posits that temperature adaptation uses existing response mechanisms, which should result in significant overlap in response. This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g. more shared "Low" transcripts) and a deficit of transcripts that shift to other response patterns or become not responsive.
